---
title: "rrn copy number assignment — example analysis"
author: "Tahiel Maccario"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

This document illustrates how to use the output of `hsp.py` to assign rrn copy
numbers to ASVs and classify them into trophic strategies. It assumes you have
already run the pipeline described in the repository README.

The inputs needed are:

| File | Description |
|------|-------------|
| `your_taxonomy.csv` | Taxonomy table with one row per ASV |
| `your_otu_table.tsv` | OTU/ASV abundance table (ASVs × samples) |
| `rrn_copies_estimated.tsv` | Output of `hsp.py` |

---

## Load data

```{r load-data}
library(dplyr)
library(ggplot2)
library(tidyr)
library(patchwork)
library(knitr)
library(kableExtra)

# ── Paths ──────────────────────────────────────────────────────────────────────
# Adjust these paths to match your project structure
tax_table  <- read.csv("your_taxonomy.csv")
otu_table  <- read.table("your_otu_table.tsv", sep = "\t", header = TRUE)
rrn_raw    <- read.table("rrn_copies_estimated.tsv", sep = "\t", header = TRUE)
```

The taxonomy table should have at minimum an `ASV` column and taxonomic rank
columns (`Phylum`, `Class`, etc.). The OTU table should have ASVs as rows and
samples as columns, with an `ASV` identifier column.

---

## Merge copy numbers into the taxonomy table

```{r merge}
# Long-format OTU table for per-sample operations
otu_long <- otu_table %>%
  pivot_longer(cols = -ASV, names_to = "sample", values_to = "abundance")

# Total abundance per ASV across all samples (used for weighted summaries)
asv_total_abundance <- otu_long %>%
  group_by(ASV) %>%
  summarise(total_abundance = sum(abundance, na.rm = TRUE), .groups = "drop")

# Join copy numbers and NSTI into the taxonomy table
taxonomy_final <- tax_table %>%
  left_join(asv_total_abundance, by = "ASV") %>%
  left_join(
    rrn_raw %>% select(
      ASV   = sequence,
      rrn   = metadata_mean_16S_rRNA_Count,
      NSTI  = metadata_NSTI
    ),
    by = "ASV"
  )
```

---

## Classify trophic strategy

Following [Klappenbach et al. (2000)](https://doi.org/10.1128/AEM.66.4.1328-1333.2000)
and [Gao & Wu (2019)](https://doi.org/10.1101/350348):

| Category | rrn copies |
|----------|-----------|
| Oligotroph | 1–2 |
| Intermediate | 3–4 |
| Copiotroph | ≥5 |

```{r classify}
taxonomy_final <- taxonomy_final %>%
  mutate(
    strategy = case_when(
      rrn <= 2              ~ "Oligotroph",
      rrn >= 5              ~ "Copiotroph",
      !is.na(rrn)           ~ "Intermediate",
      TRUE                  ~ NA_character_
    ),
    strategy = factor(strategy, levels = c("Oligotroph", "Intermediate", "Copiotroph"))
  )
```

---

## Summary tables

### ASVs per trophic strategy

```{r table-asvs}
taxonomy_final %>%
  filter(!is.na(strategy)) %>%
  count(strategy) %>%
  kable(caption = "Number of ASVs per trophic strategy",
        format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

### Total read abundance per trophic strategy

```{r table-abundance}
taxonomy_final %>%
  filter(!is.na(strategy)) %>%
  group_by(strategy) %>%
  summarise(total_reads = sum(total_abundance, na.rm = TRUE), .groups = "drop") %>%
  kable(caption = "Total read abundance per trophic strategy",
        format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

---

## Placement quality (NSTI)

The NSTI (Nearest Sequenced Taxon Index) reflects the phylogenetic distance
between each ASV and its nearest reference in rrnDB. Lower values indicate
more reliable copy number estimates.

```{r nsti-table}
taxonomy_final %>%
  filter(!is.na(NSTI)) %>%
  summarise(
    `NSTI < 0.15` = sum(NSTI < 0.15),
    `NSTI < 0.50` = sum(NSTI < 0.50),
    `NSTI < 2.00` = sum(NSTI < 2.00),
    `Total with NSTI` = n()
  ) %>%
  kable(caption = "ASVs by NSTI threshold",
        format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r nsti-hist, fig.width=6, fig.height=4}
taxonomy_final %>%
  filter(!is.na(NSTI)) %>%
  ggplot(aes(x = NSTI)) +
  geom_histogram(bins = 80, fill = "steelblue", color = "white", linewidth = 0.2) +
  geom_vline(xintercept = c(0.15, 0.50), linetype = "dashed",
             color = c("orange", "red")) +
  annotate("text", x = 0.17, y = Inf, label = "0.15", vjust = 2,
           hjust = 0, color = "orange", size = 3) +
  annotate("text", x = 0.52, y = Inf, label = "0.50", vjust = 2,
           hjust = 0, color = "red", size = 3) +
  labs(x = "NSTI", y = "Number of ASVs",
       title = "Distribution of NSTI values") +
  theme_minimal()
```

High NSTI values are expected for samples from poorly characterized environments
(e.g., freshwater systems from underrepresented regions) where rrnDB coverage is
limited.

---

## rrn copy number by phylum

```{r boxplot-phylum, fig.width=10, fig.height=7}
# Select the 15 most abundant phyla
top_phyla <- otu_long %>%
  left_join(tax_table %>% select(ASV, Phylum), by = "ASV") %>%
  filter(!is.na(Phylum)) %>%
  group_by(Phylum) %>%
  summarise(total_abundance = sum(abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_abundance)) %>%
  slice_head(n = 15) %>%
  pull(Phylum)

# Order phyla by median rrn copy number
phylum_order <- taxonomy_final %>%
  filter(Phylum %in% top_phyla, !is.na(rrn)) %>%
  group_by(Phylum) %>%
  summarise(med_rrn = median(rrn), .groups = "drop") %>%
  arrange(med_rrn) %>%
  pull(Phylum)

plot_data <- taxonomy_final %>%
  filter(Phylum %in% top_phyla, !is.na(rrn)) %>%
  mutate(Phylum = factor(Phylum, levels = phylum_order)) %>%
  arrange(total_abundance)   # most abundant points drawn on top

# Weighted mean and SD per phylum (for the annotation table)
stats <- plot_data %>%
  group_by(Phylum) %>%
  summarise(
    wmean = round(sum(rrn * total_abundance) / sum(total_abundance), 2),
    wsd   = round(sqrt(sum(total_abundance * (rrn - sum(rrn * total_abundance) /
                              sum(total_abundance))^2) / sum(total_abundance)), 2),
    .groups = "drop"
  ) %>%
  mutate(Phylum = factor(Phylum, levels = phylum_order))

tbl <- stats %>%
  pivot_longer(cols = c(wmean, wsd), names_to = "stat", values_to = "value") %>%
  mutate(value = as.character(value)) %>%
  ggplot(aes(x = stat, y = Phylum, label = value)) +
  geom_text(size = 3.2) +
  scale_x_discrete(labels = c(wmean = "", wsd = "")) +
  labs(title = "Wtd. mean   Wtd. SD") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 9),
        plot.margin = margin(5, 5, 5, 0))

bp <- ggplot(plot_data, aes(x = Phylum, y = rrn)) +
  geom_boxplot(fill = "lightblue", color = "black", outlier.shape = NA) +
  stat_summary(fun = median, geom = "crossbar",
               width = 0.6, color = "darkorange", linewidth = 0.8) +
  geom_jitter(aes(color = total_abundance, size = total_abundance),
              width = 0.2, alpha = 0.6) +
  geom_hline(yintercept = 2.5, linetype = "dashed", color = "red") +
  coord_flip() +
  scale_color_viridis_c(name = "Total ASV\nabundance (log10)", trans = "log10") +
  scale_size_continuous(trans = "log10", range = c(0.5, 3), guide = "none") +
  labs(x = "Phylum", y = "16S rRNA copy number (rrnDB)",
       title = "rrn copy number by phylum (top 15 by abundance)") +
  theme_minimal() +
  theme(plot.margin = margin(5, 0, 5, 5))

(bp | tbl) + plot_layout(widths = c(5, 2))
```

The dashed red line marks the threshold between oligotrophs (≤2 copies) and
intermediates (≥3 copies). Points are colored and sized by total ASV abundance
across all samples.

---

## References

- Klappenbach JA, Dunbar JM, Schmidt TM. 2000. rRNA operon copy number reflects ecological strategies of bacteria. *Appl Environ Microbiol* 66:1328–1333.
- Gao Y, Wu M. 2019. Free-living bacterial communities are mostly dominated by oligotrophs. *bioRxiv* doi:10.1101/350348.
- Douglas GM et al. 2020. PICRUSt2 for prediction of metagenome functions. *Nat Biotechnol* 38:685–688.
